---
title: "DAY__2"
author: "eric"
date: "2/22/2020"
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# INTRODUCTION TO R
## R operators
Variable assignment operator  <-
```{r}



y <- "HELLO"
y
```
#Arithmetic operators
##Division of numbers with operator that give an answer to one decimal place if answeris not a whole number.  /
```{r}

5/2
```
#%/% This division give answer rounded to the lower immeadiate whole number if the answer is in decimal form.
```{r}



5%/%2
```
 %% This operator gives the modulas of two numbers.
```{r}



5%%2
```


# Assigning a vector to a variable
```{r}

idea <- c(9,92,100,263,9)
mean (idea) #finding mean of values in vector idea.
sd(idea)   #finding  stardard deviation of values in vector idea.
var(idea)    #finding variance of values in vector idea.
median(idea)   #finding median of values in vector idea.
price1 <- c(100,200,NULL) # gives the  non-null values.
price1
price2 <- c(100,200,NA) # gives all values.
price2
is.na (price2) # finding missing values in vector price2.
```
# ------- intro to vectors--------
```{r}



age <- c(22,23,24,25)
names(age) <- c("cate","tom","tom","andi")
age
age["andi"]
```
# --------- intro to lists----------------- 
Assigning a list to a variable y
```{r}



y <- list(1,2,3,4)
y
```
# Combining two vectors  into a list and assigning the list to variable data.
```{r}



data <- list(names1=c("cate","tom","tom","andi"),age1=c(23,24,25,26),retired=c(T,F))
data

```
# extracting elements from data list using their indexes
```{r}
data$age1[3]
data[[1]][3]
```
# combining vectors into lists
```{r}



product1 <- c("belt","brown","red","purple")
 totalprice = c(1000,2000,3500,800)
 color = c("black","blown","red","purple")
 quantity = c(5,2,1,3)
 sales <- list(product1,totalprice ,color,quantity)
 sales
 
```
 
# Quick quiz
 get the values brown, 800, black and 5 from sales and store them in varianbles of your choice
```{r}


#gets  index 1 of the sales list  and then precisely picks  the element on index 2 of the index 1 of sales. and assigns it to variable p.
p <- sales[[1]][2]
p
#gets  index 2 of the sales data list  and then precisely picks  the element on index 4 of the index 2 of sales.and assigns it to variable t.
t <- sales[[2]] [4]
t

#gets  index 2 of the sales data list  and then precisely picks  the element on index 1 of the index 3 of sales.and assigns it to variable c.
c <- sales[[3]][1]
c


#gets  index 4 of the sales list  and then precisely picks  the element on index 1 of the index 4 of sales.and assigns it to variable t.
q <- sales[[4]][1] 
q
```
# ----------------intro to data.frames
creating a table  dataset .
```{r}



product1 <- data.frame(product1,totalprice,color,quantity)
product2 <-data.frame (product1,totalprice,color,quantity, stringsAsFactors=F)
product2

#view the elements in the product2 dataframe 

View(product2)

#view the structure of dataframes :product1 and product2
str(product1)
str(product2)

#access the element(s) in  the various indexes of product2 data frame
product2[,2]
product2[2,]
product2[[2]][1]
product2[["totalprice"]]
```
# installing and the loading package datasets
```{r}



installed.packages("datasets")
library(datasets)
```
# loading and viewing the  airquality dataset and its structure.
```{r}



data(airquality)
View(airquality)
str(airquality)
```
# summarises the air quality dataset and find its column metrics :mean ,median ,minimum and maximum values etc.
```{r}



summary(airquality)
```
# a look into the airquality dataset.
```{r}



head(airquality,15)#gives the first 15 rows in dataset airquality.
tail(airquality)# gives the last 5 rows in dataset airquality.
(View  ((tail)(airquality))) #gives the 5 last row of the dataset as a table.
is.na(airquality) # gives the missing data in dataset airquality.
sum(is.na(airquality)) #gives the sum of missing values in the dataset.
which(is.na(airquality)) #gives the exact locations of the missing values in dataset airquality.
sum(is.na(airquality$Solar.R))# gives the  sum of missing values in column solar.r of dataset airquality
sum(is.na(airquality$Ozone))# gives the  sum of missing values in column oxone  of dataset airquality.
names(airquality) #gives the column names of the dataset.
colnames(airquality) # it also gives the column names of the dataset.
dim(airquality) #give the dimmensional state  of the dataset  i.e number of columns and number of rows.
nrow(airquality) #give the number of rows in the dataset
ncol(airquality)  #give the number of columns in the dataset
which(airquality$Temp==69)  #gives the index at which in the temp column of airquality dataset the value is 69.
```

# loading and viewing the  mtcars dataset and its structure.
```{r}



mtcars
data(mtcars)
View(mtcars)
str(mtcars)
```
#  A LOOK AT DIFFERENT FEATURES OF THE DATASET.
```{r}
summary(mtcars)#summarises the air quality dataset and find its column metrics :mean ,median ,minimum and maximum values etc.
head(mtcars,15)#this gives the first 15 rows of the mtcars dataset. 
tail(mtcars,2) #this gives the last 2 rows  of the mtcars dataset.
(View  ((tail)(mtcars,5))) #this gives the last 5 rows  of the mtcars dataset as a table.
is.na(mtcars) #gives the missing values in the mtcars dataset.
sum(is.na(mtcars))#gives the sum of missing values in the mtcars dataset.
which(is.na(mtcars)) #gives the index of the missing values in the mtcars dataset.
sum(is.na(mtcars$cyl))#gives the sum of missing values in the cyl column of the mtcars dataset.
sum(is.na(mtcars$disp))#gives the sum of missing values in the disp column of the mtcars dataset.
names(mtcars) #gives the column names
colnames(mtcars)#gives the column names
dim(mtcars) # gives the dimmension of the dataset i.e number of columns and number of rows.
nrow(mtcars) # gives no of rows in the dataset
ncol(mtcars) #gives no of columns in the dataset.


View(which(mtcars$cyl==4 & mtcars$vs==0))  # returns the column name  with cyt 0f 275 and vs of 0 if available.
sum(is.na(mtcars))
mtcars[["wt"]]# gives the values in the column wt in dataset mtcars.

```
# combining vectors into a dataframe.
and accesing values in the formed data frame.
```{r}



names <- c("dau","jabber","mwendwa","wanza")
age <- c(19,2,2,49)
sales <- c(2000,1500,1800,3000)
month <- c("nov","dec","oct","nov")
data.frame(names,age,sales,month)
customer_details <- data.frame(names,age,sales,month)
customer_details[,3]  
```
# ANOTHER LOOK AT DATASET MTCARS
```{r}



data("mtcars") #loading the dataset.
?mtcars # help in deep understanding of the dataset.
mtcars[,1]# gives the values of the values in the  first column of mtcars dataset.
row.names(mtcars) #gives the rownames.
colnames(mtcars)# gives the column names.
colnames(mtcars)[colnames(mtcars)=="cyl"] #searches the value where the column name is cyl.
```
# renaming columns
and setting the new column names as factors.
```{r}
colnames(mtcars)[colnames(mtcars)=="cyl"] <- "cyla"

mtcars[1:10,]
head(mtcars$vs,10)
str(mtcars)
View(mtcars)
mtcars$vs <- as.factor(mtcars$vs)
str(mtcars)

str(mtcars)
colnames(mtcars)<-c("name","genus","order","vore","sleep_total","qw","er","ty","ui","as","df")
mtcars_new <- mtcars[,colnames(mtcars)]
View(mtcars_new )
table(mtcars$genus)
rownames(mtcars_new)[row.names(mtcars_new)=="Valiant"] <- "mac x"
(row.names(mtcars_new))
as.factor(mtcars_new)
```
# dates in r  -----
loading library lubridate
```{r}



library(lubridate)
```
#DATES IN R --
  Create  data and time values and assign them to variable leo and then view structure of variable leo. shows they are in character form.
```{r}


 
leo <- ("21-01-2020 09:56:03")
str(leo)
```
# initialize the dates /time function in use with variable leo and 
assign it to reassign the variable leo to variable leol
view structure of variable leol
```{r}



leol <- dmy_hms(leo)
str(leol)
```
# using the dates function  assign leol values  to a new precise variable and then view them but in number for instead of specific month/day name.
```{r}



leol_second <- second(leol)
leol_second
leol_minute <- minute(leol)
leol_minute
leol_hour <- hour(leol)
leol_hour
leol_month <- month(leol)
leol_month
leol_year <- year(leol)
leol_year
```
# label =T ,gives the days /month name 
abbr=T ,gives the month/day name in a abbreviated form,;for instance monday given as mon ,december as dec.
```{r}



leol_day <- wday(leol,label=T,abbr=T)
leol_day
leol_month <- month(leol,label=T,abbr=T)
leol_month
```

#-------r  mini project
Importing data from file "amazon_kenya_data"
and loading package tidyverse to facilitate piping of the data.
```{r}
library(tidyverse)
Amazon <- read.csv("E:/Documents/class/amazon_kenya_data.csv",stringsAsFactors = F)

```
# A look at dataset amazon
```{r}



str(Amazon) #checking the structure of amazon dataset.
summary(Amazon) ##summarises the air quality dataset and find its column metrics :mean ,median ,minimum and maximum values etc.
head(Amazon,15) # this gives the first 15 rows of the amazon dataset.
View(Amazon) #to check the amazon dataset table .
table(Amazon$NAIVAS.BRANCH)#to check contents and frequency of values of the naivas branch of the amazon dataset table .
table(Amazon$DEPARTMENT)#to check contents and frequency of values of the department column of the amazon dataset table .
table(Amazon$CATEGORY)#to check contents and frequency of values of the category column of the amazon dataset table .
dim(Amazon) # to check the dimmension ;no of rows and columns in the dataset.
colnames(Amazon) # to check the number and names of columns in the amazon dataset.
rownames(Amazon)# to check the number of  rows in the amazon dataset.
```
# Creating a new column total price in the amazon dataset.
```{r}



Amazon$TOTAL_PRICE <- (Amazon$ITEM.PRICE*Amazon$QUANTITY.PURCHASED)
```
#Extracting the number of item in category fat and oils while showing their total price
```{r}



FAT_oils <- Amazon %>%
  select(CATEGORY,TOTAL_PRICE)%>% 
  filter(CATEGORY=="FATS & OILS") 
table(FAT_oils)
```
#Extracting the number of item in category  beverage while showing their total price.
```{r}



BEVERANGES <- (Amazon)%>% 
  select(CATEGORY,TOTAL_PRICE)%>% 
  filter(CATEGORY=="BEVERAGE") %>% group_by(CATEGORY) %>% summarise(sum(TOTAL_PRICE))
table(BEVERANGES)
```
#DEPARTMENT PERFORMANCE
```{r}



Department_performance <- Amazon %>% 
  select(DEPARTMENT, TOTAL_PRICE) %>%
  group_by(DEPARTMENT)%>%
  summarise(TOTAL_PRICE = sum(TOTAL_PRICE))
```
# Names of departments in the amazon dataset.
```{r}



names(table(Amazon$DEPARTMENT)) 
```
 #  DATA ANALYSIS
 CATEGORY_PERFORMANCE
```{r}


 
CATEGORY_PERFORMANCE <- Amazon%>% select(CATEGORY,TOTAL_PRICE)%>%
  group_by(CATEGORY)%>%
  summarise(TOTAL_PRICE=sum(TOTAL_PRICE))
```
# ITEM PERFORMANCE
```{r}



ITEM_PERFORMANCE <- Amazon%>% select(ITEM.DESCRIPTION,TOTAL_PRICE)%>%
  group_by(ITEM.DESCRIPTION)%>%
  summarise(TOTAL_PRICE=sum(TOTAL_PRICE))
```
# BRAND PERFORMANCE
```{r}



BRAND_PERFORMANCE <- Amazon%>% select(BRAND,TOTAL_PRICE)%>%
  group_by(BRAND)%>%
  summarise(TOTAL_PRICE=sum(TOTAL_PRICE))
```
# BRANCH PERFORMANCE
```{r}



BRANCH_PERFORMANCE <- Amazon%>% select(NAIVAS.BRANCH,TOTAL_PRICE)%>%
  group_by(NAIVAS.BRANCH)%>%
  summarise(TOTAL_PRICE=mean(TOTAL_PRICE))
```
# CUSTOMER PERFORMANCE IN   NUMBER OF TRANSACTIONS MADE
```{r}



CUSTOMER_PERFORMANCE <- Amazon%>% select(CUSTOMER.NAME,CUSTOMER.ID,TOTAL_PRICE)%>%
  group_by(CUSTOMER.NAME )%>%
  summarise (count=n())
```
# CUSTOMER PERFORMANCE IN  RELATION TO TOTALS
```{r}



CUSTOMER_PERFORMANCE <- Amazon%>% select(CUSTOMER.NAME,CUSTOMER.ID,TOTAL_PRICE)%>%
  group_by(CUSTOMER.NAME )%>%
  summarise (sum(TOTAL_PRICE))
```
# CUSTOMER FREQUENCY
```{r}



CUSTOMER_FREQUENCY <- Amazon%>% select(CUSTOMER.NAME,CUSTOMER.ID,TOTAL_PRICE)%>%
  group_by(CUSTOMER.NAME )%>%
  summarise (count=n())

```





