---
title: "SUPERMARKET_ANALSIS"
author: "eric"
date: "2/19/2020"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



#This is a supermarket dataset.
##To facilitate better bussiness decision making data analysis is
##a necessity.
## Calling necessary libraries
```{r}



library(tidyverse)
library(ggplot2)
library(gganimate)
```
#Importing data
```{r}



SUPERMARKET <- read.csv("Supermarket Dataset.csv",stringsAsFactors = F)
```
#Cleaning data.
```{r}
SUPERMARKET[296,"date"] <- "27/06/17"
SUPERMARKET[297,"date"]
#-----------------------
index <- which(SUPERMARKET$date =="28-06-17")
SUPERMARKET[index,"date"] <- "28/06/17"
#-----------------------
which(SUPERMARKET$date == "17?06")
table(SUPERMARKET$date)

#separating columns date   and time
SUPERMARKET <- SUPERMARKET%>% separate(col=date ,into = c("day","month","year"),sep = "/")
SUPERMARKET <- SUPERMARKET%>% separate(col=time ,into = c("hour","minute"),sep = ":")
```
#checking indexes and removing anomalies data.


```{r}
table(SUPERMARKET$day)

index <- which(SUPERMARKET$day =="17?06" |SUPERMARKET$day =="2305")
SUPERMARKET <- SUPERMARKET[-c(768),]
  
```
#ANALYSIS
SUPERMARKETS PERFORMANCE
```{r}



supermarket_performance <- SUPERMARKET%>%
  select(total,supermarket)%>%
  group_by(supermarket)%>%
  summarise(total = sum(total)) %>% 
  arrange(desc(total))
#It has been noted above that NAKUMATT SUPERMARKETS ARE THE BEST IN FINANCIAL PERFOMANCE.
#However kassmart supermarkets and foodplus lugs behind in the performance list.
```


#BUSINESS LOCATION PERFORMANCE  ANALYSIS
```{r}



LOCATION_PERFORMANCE<- SUPERMARKET%>%
  select(total,location,supermarket)%>%
  group_by(location)%>%
  summarise(total = sum(total))%>%
  arrange(desc(total))
#Location performance analysis proofs that
#supermarket business in the cbd is the best.
#However karasha  location and kericho town  location  have the worst business performance.
```

#MODE OF PAYMENTS PERFORMANCE  ANALYSIS
```{r}


mode_of_PAYMENT_PERFORMANCE <- SUPERMARKET%>%
  select(total,type,supermarket)%>%
  group_by(type)%>%
  summarise(total = sum(total)) %>%
  arrange(desc(total))
#From the above mode of payment analysis its crear that most customers still
#prefer cash payments.Payments using the voucher is
#the most unpopular mode of payment.
```

#DAYS_OF_THE_MONTH_PERFORMANCE ANALYSIS
```{r}


 
DAYS_OF_THE_MONTH_PERFORMANCE <-SUPERMARKET%>%
  select(total,day,month, supermarket)%>%
  group_by(day)%>%
  summarise(total = sum(total)) %>% 
  ungroup() %>%
  mutate(day = as.integer(day))%>%
  group_by(day) %>%
  filter(!day == 	2305)%>%
  summarise(total = sum(total)) %>%
  arrange(desc(total))
  na.omit
   


#With the proof of the above analysis its noted that sales
#are optimum on the 23 rd day of the month and then decrease towards
#the mid-month of of the succeding month.
```

#PERFORMANCE_PER_MONTHS_OF_THE_YEAR ANALYSIS
```{r}



PERFORMANCE_PER_MONTHS_OF_THE_YEAR <- SUPERMARKET%>%
  select(total,month,supermarket)%>%
  mutate(month = as.integer(month))%>%
  group_by(month)%>%
  summarise(total = sum(total)) %>%
  filter(!month == 2017)%>%
arrange(desc(total))
  
#From the above months of the year  financial analysis it can be induced that  best performing
#month is month ,may with  a total value of sales being  of 114102.90 ,while month feb
#is the worst performing month with
#a total value of sales being 90.00 .

```

#DAY_TYPE_performance
```{r}



DAY_TYPE_performance <- SUPERMARKET%>%
  select(total,day_type)%>%
  group_by(day_type)%>%
  summarise(meantotal=  mean(total))%>%
arrange(desc(meantotal))
#On average weekends supermarkets seemed to be performing better than on weekdays.
#This suggests that many customers shop during weekends.
```

#24 HOUR BUSINESS OPERATION ANALYSIS
```{r}



supermarkets_OPEN_24_HR_VS_NOT <- SUPERMARKET%>%
  select(total,open_24_hr,supermarket)%>%
  group_by(open_24_hr)%>%
  summarise(count= n())%>%
arrange(desc(count))
supermarket_OPEN_24_HR_performance <- SUPERMARKET%>%
  select(total,open_24_hr)%>%
  filter(open_24_hr=="yes")%>%
  group_by(open_24_hr)%>%
  summarise(mean=mean(total))%>%
arrange(desc(mean))
supermarket_not_OPEN_24_HR_performance <- SUPERMARKET%>%
    select(total,open_24_hr)%>%
    filter(open_24_hr=="no")%>%
    group_by(open_24_hr)%>%
    summarise(mean=mean(total))%>%
arrange(desc(mean))
#In the above analysis it is proved that the number non-24-hr supermarkets is greater than the 
#number of 24-hr-supermarkets.
#Moreover the average performance of non-24-hr supermarkets is better than the one
#for 24-hr-based supermarkets.
```


#FOOD_SALES_ANALYSIS
```{r}



BOUGHT_FOOD <- SUPERMARKET%>%
  select(total,food,location,supermarket)%>%
  filter(food=="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
arrange(desc(count))
#It was noted that food sales were highest in the cbd location with a total sales
#transactions being 349.
```
#FOOD_AND_SNACK_COMBINATION_SALES_ANALYSIS
```{r}



BOUGHT_FOOD_AND_SNACK <- SUPERMARKET%>%
  select(total,food,snack,location,supermarket)%>%
  filter(food=="yes",snack=="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#It was noted,from the above analysis that food and snacks combination sales were best in the 
#cbd location with a total number of  the transactions being 82.
```
#FOOD_AND_SNACK_BEVERAGE_COMBINATION_SALES_ANALYSIS
```{r}



BOUGHT_FOOD_AND_SNACK_AND_BEVERAGE <- SUPERMARKET%>%
  select(total,food,snack,beverage,location,supermarket)%>%
  filter(food=="yes",snack=="yes",beverage=="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#From the above analysis it was clear that the food and snack and beverage (together) transactions
#were most in the cbd location with a total number of transactions being 31.
```


#BEVERAGE_and_SNACKS_SALES_ANALYSIS
```{r}



BOUGHT_BEVERAGE_and_snacks <- SUPERMARKET%>%
  select(total,food,beverage,snack,location,supermarket)%>%
  filter(beverage=="yes",snack=="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#It was noted,from the above analysis that beverages and snacks combination (together) sales were best in the 
#cbd location with a total number of  the transactions being 81.

```


#SNACK__SALES_ANALYSIS
```{r}



BOUGHT_SNACK <- SUPERMARKET%>%
  select(total,food,snack,location,supermarket)%>%
  filter(snack=="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#It was noted,from the above analysis that  consumables sales were best in the 
#cbd location with a total number of  the transactions being 288.



```


#BEVERAGE_SALES_ANALYSIS
```{r}



BOUGHT_BEVERAGE <- SUPERMARKET%>%
  select(total,month,beverage,location)%>%
  filter(beverage=="yes")%>%
  mutate(month=as.integer(month))%>%
  group_by(location,month)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#THE above beverage sales analysis proved that beverage business was best in the cbd location
#with beverage total transactions being 289.
```


#CONSUMABLES SALES_ANALYSIS
```{r}


BOUGHT_consumables <- SUPERMARKET%>%
  select(total,food,consumables,location,supermarket)%>%
  filter(consumables  =="yes")%>%
  group_by(location)%>%
  summarise(count= n())%>%
  arrange(desc(count))
#It was noted,from the above analysis that  snacks sales were best in the 
#cbd location with  a total number of  179 consumables transactions 
#followed by ongata rongai location then third donholm 
#location .
```



#MALLS_VS_NONMALLS_PERFORMANCE ANALYSIS
```{r}



MALLS_VS_NONMALLS_PERFORMANCE <- SUPERMARKET%>%
  select(total,mall,supermarket)%>%
  group_by(mall)%>%
  summarise(mean=mean(total))%>%
  arrange(desc(mean))
#From the above analysis it has been noted that supermarket
#in malls perform better than those in the open market.
```

#NUMBER_OF_MALLS_IN_A_LOCATION ANALYSIS
In the above data analysis it is creally proved that the number of malls are 
more in the locations (juction , followed by donholm then in kilimani).
This further proves that the level of business performance in this respective 
locations is optimal compared to other locations.
 
```{r}



NO_OF_MALLS_IN_A_LOCATION<- SUPERMARKET%>%
  select(mall,total,location)%>%
  filter(mall=="yes")%>%
  group_by(location)%>%
  summarise (count= n())%>%
  arrange(desc(count))

```


#CREATING A DATASET YEAR AND DATA CLEANING IN COLUMN MONTH
```{r}



YEAR<- SUPERMARKET%>%
select(year,month,total)%>%
  mutate(month = as.integer(month))%>%
group_by(year,month)%>%
summarise (total = sum(total))%>%
arrange(desc(total))%>% na.omit()
YEAR
str(YEAR$year)

index <- which(YEAR$year == "201"|YEAR$year == "2013"|YEAR$year == "2015"|
        YEAR$year == "2016")
index <- as.vector(index)

YEAR[index,1] <-  "2017"
YEAR[c(12,19,24,25,26,27,28,29,30,32,33),1] <-  "2018"


YEAR <- YEAR %>% group_by(year, month) %>% summarise(total = sum(total))
```

  
  
  
   #SUPERMARKET DATASET  VISUALIZATION
 
 
 
##THE FREQUENCY OF THE DIFFERENT TOTALS OF SHOPPING
 ###using a histogram
```{r}
ggplot(SUPERMARKET,aes(x=total ))+
   geom_histogram(position="identity",bins=60,fill = "green")+
   scale_x_log10()+
   labs(title = "THE FREQUENCY OF THE DIFFERENT TOTALS OF SHOPPING",x="TOTAL MONEY",y="count")+
   theme(panel.background =element_rect(fill="darkolivegreen"))+
   theme(plot.title  = element_text(colour ="darkgreen",face = "bold",hjust = 0.5))
  
# scale_fill_manual(values = c("green","light green"))
 
```
 
 ##THE FREQUENCY OF THE DIFFERENT TOTALS OF SHOPPING
 ###using a frequency-polygon
```{r}


ggplot(SUPERMARKET,aes(x=total ))+
   geom_freqpoly(size=1,position="identity",bins=60,col="green")+
   scale_x_log10()+
   labs(title = "THE FREQUENCY OF THE DIFFERENT TOTALS OF SHOPPING",x="TOTAL MONEY",y="count")+
   theme(panel.background =element_rect(fill="mediumseagreen"))+
   theme(plot.title  = element_text(colour ="darkgreen",face = "bold",hjust = 0.5))
 
   #It can be noted that most of the prices of the goods bought are in the range of 50 to 1000
 #shilings.
 
```
 
 #VISUALISATION OF CORRELATION OF TOTAL MONEY VS NO OF ITEMS IN RELATION TO FOODS 
 #AND NON-FOODS ITEMS
```{r}

  ggplot(SUPERMARKET, aes(y=total,x=no_of_items,col=food))+
   geom_point(size = 3 )+geom_smooth(method="lm",se=F)+
   labs(title = "TOTAL MONEY VS NO OF ITEMS and food / non-food items",y="TOTAL MONEY",x="NO OF ITEMS")+
   theme(axis.text.x = element_text(angle = 45 ,vjust = 0.5))+
  theme(plot.title = element_text(hjust = 0.5 ,face = "bold"))+
   theme(panel.background  = element_rect(fill = "lightgreen"))

  # It can be induced that  food items are more relatively expensive than  non-food items
 #CORRELATION
```
  
  
  #TOTAL MONEY EARNED IN RELATION TO MODE OF PAYMENT USED/TIME_TYPE
```{r}


  
  ggplot(SUPERMARKET,aes(x=(type),y=total,col=as.factor(time_type)))+
    geom_boxplot()+
    scale_y_log10()+
  labs(title = "TOTAL MONEY EARNED IN RELATION TO MODE OF PAYMENT USED/TIME_TYPE",x="MODE OF PAYMENT",y="TOTAL MONEY EARNED")+
   labs(col="TIME_TYPE") +
 theme(plot.title = element_text(hjust=0.5,colour = "darkgreen",face="bold"))+ 
  theme(panel.background = element_rect(fill="green"))+
  theme(legend.background = element_rect(fill="palegreen"))+
  theme(axis.text.x=element_text(angle = 45))
  
  #From the above its clear that the credit card is highly used during the night  . 
  #voucher is the least  performing mode of payment and its used during the evening.
  
```
  
  #TOTAL MONEY EARNED IN RELATION TO NO OF ITEMS BOUGHT/TIME_TYPE
```{r}


  
  
  ggplot(SUPERMARKET,aes(x=(no_of_items),y=total,col=as.factor(time_type)))+
    geom_smooth(method="lm",se=F)+
    #scale_y_log10()+
    labs(title = "TOTAL MONEY EARNED IN RELATION TO NO OF ITEMS BOUGHT/TIME_TYPE",x="NO OF ITEMS BOUGHT",y="TOTAL MONEY EARNED")+
    labs(col="TIME_TYPE") +
    theme(plot.title = element_text(hjust=0.5,colour = "darkgreen",face="bold"))+ 
    theme(panel.background = element_rect(fill="limegreen"))+
    theme(legend.background = element_rect(fill="palegreen"))+
    theme(axis.text.x=element_text(angle = 45))
  
  #The above shows that more expensive items are bought during the morning bussiness hours and the 
  #overal business financial performance is highest in the morning transactions.
```
  
  
  
 #MONTHS OF THE YEAR PERFOMANCE
```{r}


 
ggplot(YEAR,aes(y=total,x=factor(month)))+
  geom_bar(stat="identity",fill="forestgreen")+
 labs(x="month",y="total",title = "MONTHS OF THE YEAR PERFORMANCE")+
  theme(panel.background = element_rect(fill = "limegreen"))
#It was noted that the most productive MONTHS of the year are the 5th and  6th  month.




```

 #DAYS OF THE MONTH PERFORMANCE
```{r}


 
ggplot(DAYS_OF_THE_MONTH_PERFORMANCE,aes(x=factor(day),y=total))+
  geom_bar(stat="identity",fill="darkolivegreen")+
  labs(title = "DAYS OF THE MONTH PERFORMANCE",x="day of the month",y="totals MONEY")+
theme(plot.title = element_text(hjust = 0.5,face = "bold",colour = "darkgreen"))+
  theme(panel.background = element_rect(fill="lightgreen"))
#It was noted that the most productive days of the month are the 22nd and  23 rd day.

```
#SALES OF BEVERAGE DURING THE YEAR
```{r}



ggplot(BOUGHT_BEVERAGE,aes(x=factor(month),y=count))+
  geom_bar(stat = "identity",fill="limegreen")+
labs(title = "SALES OF BEVERAGE DURING THE YEAR",x="MONTH OF THE YEAR",y="NO OF BAVERAGES")+
  theme(plot.title = element_text(hjust = 0.5,colour ="green",face = "bold"))+
  theme(panel.background = element_rect(fill="palegreen"))

 #it can be noticed that the highest no of beverages are bought during the 5th month.

```

#NO OF MALLS IN A LOCATION 
##using a  bar graph
```{r}
 ggplot(NO_OF_MALLS_IN_A_LOCATION, aes(reorder(x=as.factor(location),count), y = (count)))+
  geom_bar(stat = "identity", position = "dodge",fill="lawngreen")+
  labs(title = "NO OF MALLS IN A LOCATION",x="LOCATION",y="COUNT OF MALLS")+
  theme(axis.text.x = element_text(angle = 5,colour = "limegreen" ,vjust = 0.5))+
theme(plot.title = element_text(hjust = 0.5 ,colour="green",face = "bold"))+
theme(panel.background = element_rect(fill = "forestgreen"))

```
#NO OF MALLS IN A LOCATION
##USING A SCATTER PLOT
```{r}



ggplot(NO_OF_MALLS_IN_A_LOCATION,aes(x=location,y=count))+
  geom_point(size=5,colour="blue")+
  
  labs(title = "NO OF MALLS IN A LoCATION ",x="LOCATION",Y="NO OF MALLS")+
theme(plot.title = element_text(hjust = 0.5,colour = "cyan",face = "bold"))+
  theme(panel.background = element_rect(fill = "green"))+
  theme(axis.text.x = element_text(angle = 45,colour = "purple"))+
theme(plot.background = element_rect(fill="palegreen"))
#IT has been noted the junction location has the largest no of malls whereas likoni 
#location has the least no of malls.

```


 